---
title: "ESM 232 Hunting Assignment"
author: "Maxwell Pepperdine"
format: html
editor: visual
---

## Load packages

```{r}
rm(list=ls())
```

```{r}
library(here)
library(lhs)
library(epiR)
library(tidyverse)
library(deSolve)
```

## Overview

...

## Part 1

#### Build the hunting model (e.g add hunting to the `lotvmodK.R`)

**Hints:** - make sure that we don't hunt more prey than exist. - to ensure this, we might also add a minimum prey population input that must be met before hunting is allowed (only turn on hunting above this value). - Note: we can make this as simple or as complex as you would like. - We could represent hunting in a way that is similar to "harvesting" in the last assignment.

**Discussion Notes:** - Think about how we can represent hunting in a way that is similar to "harvesting" in the last assignment (only allow it when populations are above a certain value).

## Part 2

#### Overview

**References:**

-   `Lecture12_discussion.qmd` from `ESM232_EDS230_Examples` repository

**Discussion Notes:** - To define stability, it can be as simple as maintaining a population above some value 50 years into the future

Our goal is to explore how different hunting levels and different minimum prey populations (before hunting is allowed) are likely to effect the stability of the populations of both predator and prey. To do this, we will define a stability metric which we can use to estimate sustainable hunting levels, play around with different parameter sets, and then use this exploration to recommend a hunting target that will be sustainable (e.g leave you with a stable prey and predator population).

#### Defining/quantifying stability



















